Multilevel spin models in cold atomic systems

### 2021-04-05 sun_phases/spin_model_paper/main.tex
choose 'multi-cat' states to simulate

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}

@@

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as $H_{\t{int}} \to H_{\t{int}}^{(\phi)}$, where
\begin{align}
  H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $s\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ ($J$ large compared to $U$) while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q}.
  \label{eq:H_spin}
\end{align}

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, when $n$ is even we can prepare an $N$-fold product of the state
\begin{align}
  e^{-\i\frac{\pi}{2}\p{s_\y+s_\y^2}} \ket{s} \simeq \ket{s} + \ket{-s},
\end{align}
where $\simeq$ denotes equality up to a global phase and normalization factor.
This state has a vanishing mean spin vector, $\bk{s_\x}=\bk{s_\y}=\bk{s_\z}=0$, but variances $\bk{s_\x^2}=\bk{s_\y^2}=s/2$ and $\bk{s_\z^2}=s^2$.

@@

We now return to the spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin}, which is valid in the weak-SOC limit $s\phi\ll1$, and henceforth work exclusively in the ``gauged frame'' of $H_{\t{spin}}$ and the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
At the mean-field (MF) level, up to constant energy shifts the undriven spin Hamiltonian becomes
\begin{align}
  H_\MF = \sum_q\sp{-U\bk{\bar{\v s}}\c\v s_q
    + J\phi \sum_q \sin\p{q} s_{\z,q}}.
  \label{eq:H_MF}
\end{align}
where $\bar{\v s}\equiv\frac1N\sum_q\v s_q$ is the average spin matrix.
Fixing the atom number $N$, the spin Hamiltonian has one free parameter: the reduced field $J\phi/U$ that determines the relative strength of the single-particle and interaction terms.
One should therefore expect distinct dynamical behaviors when $J\phi/U\ll1$, in which case strong spin-locking interactions should give rise to a long-range ordered phase, as opposed to $J\phi/U\gg1$, in which case long-range order should be destroyed by the strong inhomogeneous magnetic field.

To investigate these behaviors quantitatively, we examine time-averaged observables of the form
\begin{align}
  \bbk{\O}_\MF = \lim_{T\to\infty} \f1T \int_0^T \dd t \bk{\O\p{t}}_\MF,
\end{align}
where $\bk{\O\p{t}}_\MF$ is the mean-field value of observable $\O$ at time $t$ (see Appendices \ref{sec:MFT} and \ref{sec:bosons} for details about our mean-field simulations).
Specifically, we consider the time-averaged magnetization
\begin{align}
  m_\MF \equiv \abs{\bbk{\vec\sigma}_\MF},
  &&
  \vec\sigma \equiv \f1{Ns} \times \p{S_\x,S_\y,S_\z},
\end{align}
where $S_\alpha \equiv \sum_q s_{\alpha,q}$, and the time-averaged interaction energy
\begin{align}
  \bbk{\ss}_\MF = \f1{N^2} \times \bbk{\v S\c\v S}_\MF.
\end{align}
By design, these quantities are non-negative and normalized to a maximal value of one, independent of the system size $N$ or spin dimension $n$.
In the remainder of this section we will assume that $n$ is even, both for the sake of experimental relevance (most relevant atomic nuclei are fermionic) and to avoid complications from parity effects.

\begin{figure}
\centering
\includegraphics{mean_mag_int.pdf}
\caption{
Time-averaged magnetization $m_\MF$ and interaction energy $\bbk{\ss}_\MF$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket\x^{\otimes N}$ up to a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $\bbk{\bar\sigma_\x}_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_mag_int}
\end{figure}

Figure \ref{fig:mean_mag_int} shows the time-averages of the magnetization $m_\MF$ and interaction energy $\bbk{\ss}_\MF$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket\x^{\otimes N}$, where
\begin{align}
  \ket\x \equiv e^{-\i\frac{\pi}{2}s_\y} \ket{s}
  = \f1{2^s} \sum_\mu { 2s \choose s+\mu }^{1/2} \ket{\mu}.
\end{align}
Here ${ m \choose k }$ is a binomial coefficient.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero magnetization $m_\MF$ and an interaction energy $\bbk{\ss}_\MF$ that asymptotically approaches its maximal value as $J\phi/U\to0^+$.
The disordered phase has no (time-averaged) magnetization, $m_\MF=0$, but an interaction energy $\bbk{\ss}_\MF$ that nonetheless indicates persistent nontrivial inter-spin correlations when $n>2$.
These nontrivial inter-spin correlations vanish as $J\phi/U\to\infty$, in which case $\bbk{\ss}_\MF$ approaches the minimal value allowed by conservation laws (clarified below).
By minimizing the reduced field $J\phi/U$ for which $m_\MF=0$, we numerically find that the transition between ordered and disordered phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields}).

@@

In the weak-field limit $J\phi/U\to0^+$, the spin-locking $\v S\c\v S$ interactions of the spin Hamiltonian $H_{\t{spin}}$ energetically restrict dynamics to the permutationally-symmetric (PS) manifold.
At second order in perturbation theory, the effective spin Hamiltonian within the PS manifold becomes (see Appendix \ref{sec:pert_theory})
\begin{align}
  H_{\t{spin}}^{\t{eff}}
  = \f{\p{J\phi}^2}{\p{N-1}U}
  \times \sp{S_\z^2 - N\sum_q s_{\z,q}^2},
\end{align}
whose mean-field version is
\begin{align}
  H_{\t{MFT}}^{\t{eff}} = -\f{\p{J\phi}^2}{U} \sum_q s_{\z,q}^2,
\end{align}
where we have used the fact that the axial magnetizations $\bk{s_{\z,q}}=0$ are conserved.
The weak-field effective Hamiltonian preserves permutational symmetry, so $\bbk{\ss}_\MF\to1$ as $J\phi/U\to0^+$.
In addition, invariance of the initial state $\ket\x^{\otimes N}$ and the effective Hamiltonian $H_{\t{spin}}^{\t{eff}}$ under rotation by $e^{-\i\pi S_\x}$ implies that
\begin{align}
  \bk{S_\y} = \bk{e^{\i\pi S_\x} S_\y e^{-\i\pi S_\x}} = -\bk{S_\y}
\end{align}
at all times, so $\bk{S_\y}=0$.
Together with permutational symmetry, the fact that $\bk{S_\y}=\bk{S_\z}=0$ implies that the magnetization $m_\MF$ is determined by the time-average of $\sigma_\x\equiv s_\x/s$ for a single spin:
\begin{align}
  \lim_{J\phi/U\to0^+} m_\MF
  = \abs{\lim_{T\to\infty} \f1T \int_0^T \dd\tau
  \bk{\x|\sigma_\x\p{\tau}|\x}},
\end{align}
where
\begin{align}
  \sigma_\x\p{\tau} = e^{\i\tau s_\z^2} \sigma_\x e^{-\i\tau s_\z^2}.
\end{align}
By mapping the Hilbert space of an $n$-level spin onto the PS (Dicke) manifold of $n-1$ spin-$\frac12$ particles, we can adapt exact analytical results for the Ising model \cite{foss-feig2013nonequilibrium}\footnote{See Appendix K of Ref.~\cite{perlin2020shorttime} for a simpler adaptation of the analytics in Ref.~\cite{foss-feig2013nonequilibrium} to the one-axis twisting model $H_{\t{OAT}}=\chi s_\z^2$.} to find that
\begin{align}
  \bk{\x|\sigma_\x\p{\tau}|\x} = \p{\cos\tau}^{n-2},
\end{align}
so for even $n$
\begin{align}
  \lim_{J\phi/U\to0^+} m_\MF
  = \abs{\f1{2\pi} \int_0^{2\pi} \dd\tau \p{\cos\tau}^{n-2}}
  = \gamma\p{\f{n}{2}}.
\end{align}

>>>>

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}

@@

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as $H_{\t{int}} \to H_{\t{int}}^{(\phi)}$, where
\begin{align}
  H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $s\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ ($J$ large compared to $U$) while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q},
  \label{eq:H_spin}
\end{align}
consisting of a spin-locking $\v S\c\v S$ term that energetically favors permutational symmetry, and an inhomogeneous magnetic field that causes inter-spin dephasing.

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, when $n$ is even we can prepare an $N$-fold product of the state
\begin{align}
  \frac{\ket{s} + \ket{-s}}{\sqrt{2}}
  \simeq e^{-\i\frac{\pi}{2}\p{s_\y + s_\y^2}} \ket{s},
\end{align}
where $\simeq$ denotes equality up to a global phase.
This state has a vanishing mean spin vector, $\bk{s_\x}=\bk{s_\y}=\bk{s_\z}=0$, but variances $\bk{s_\x^2}=\bk{s_\y^2}=s/2$ and $\bk{s_\z^2}=s^2$.

@@

We now return to the spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin}, which is valid in the weak-SOC limit $s\phi\ll1$, and henceforth work exclusively in the ``gauged frame'' of $H_{\t{spin}}$ and the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
At the mean-field (MF) level, up to constant energy shifts the undriven spin Hamiltonian becomes
\begin{align}
  H_\MF = \sum_q\sp{-U\bk{\bar{\v s}}\c\v s_q
    + J\phi \sum_q \sin\p{q} s_{\z,q}}.
  \label{eq:H_MF}
\end{align}
where $\bar{\v s}\equiv\frac1N\sum_q\v s_q$ is the average spin matrix.
We assume that all momenta $q\in\ZZ_N\times 2\pi/N$ are occupied.
Fixing the atom number $N$, the spin Hamiltonian has one free parameter: the reduced field $J\phi/U$ that determines the relative strength of the single-particle and interaction terms.
One should therefore expect distinct dynamical behaviors when $J\phi/U\ll1$, in which case strong spin-locking interactions should give rise to a long-range ordered phase, as opposed to $J\phi/U\gg1$, in which case long-range order should be destroyed by the strong inhomogeneous magnetic field.

To investigate these behaviors quantitatively, we examine time-averaged observables of the form
\begin{align}
  \bbk{\O}_\MF = \lim_{T\to\infty} \f1T \int_0^T \dd t \bk{\O\p{t}}_\MF,
\end{align}
where $\bk{\O\p{t}}_\MF$ is the mean-field value of observable $\O$ at time $t$ (see Appendices \ref{sec:MFT} and \ref{sec:bosons} for details about our mean-field simulations).
Specifically, we consider the time-averaged magnetization
\begin{align}
  m_\MF \equiv \abs{\bbk{\vec\sigma}_\MF},
  &&
  \vec\sigma \equiv \f1{Ns} \times \p{S_\x,S_\y,S_\z},
\end{align}
where $S_\alpha \equiv \sum_q s_{\alpha,q}$, and the time-averaged interaction energy
\begin{align}
  \bbk{\ss}_\MF = \f1{N^2} \times \bbk{\v S\c\v S}_\MF.
\end{align}
By design, these non-negative quantities are normalized to a maximal value of one, independent of the system size $N$ or spin dimension $n$.
In the remainder of this section we will assume that $n$ is even, both for the sake of experimental relevance (most relevant atomic nuclei are fermionic) and to avoid complications from parity effects.

\subsection{Spin-polarized initial state}

\begin{figure}
\centering
\includegraphics{mean_mag_int.pdf}
\caption{
Time-averaged magnetization $m_\MF$ and interaction energy $\bbk{\ss}_\MF$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket\x^{\otimes N}$ for a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $m_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_mag_int}
\end{figure}

Figure \ref{fig:mean_mag_int} shows the time-averages of the magnetization $m_\MF$ and interaction energy $\bbk{\ss}_\MF$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket\x^{\otimes N}$, where
\begin{align}
  \ket\x \equiv e^{-\i\frac{\pi}{2}s_\y} \ket{s}
  = \f1{2^s} \sum_\mu { 2s \choose s+\mu }^{1/2} \ket{\mu}.
\end{align}
Here ${ m \choose k }$ is a binomial coefficient.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero magnetization $m_\MF$ and an interaction energy $\bbk{\ss}_\MF$ that asymptotically approach their maximal values as $J\phi/U\to0^+$.
The disordered phase has no (time-averaged) magnetization, $m_\MF=0$, but the interaction energy $\bbk{\ss}_\MF$ nonetheless indicates persistent nontrivial inter-spin correlations when $n>2$.
These nontrivial correlations vanish as $J\phi/U\to\infty$, in which case $\bbk{\ss}_\MF$ approaches the minimal value allowed by conservation laws (clarified below).
By minimizing the reduced field $J\phi/U$ for which $m_\MF=0$, we numerically find that the transition between ordered and disordered phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields}).

@@

In the weak-field limit $J\phi/U\to0^+$, the spin-locking $\v S\c\v S$ interactions of the spin Hamiltonian $H_{\t{spin}}$ energetically restrict dynamics to the permutationally-symmetric (PS) manifold.
At second order in perturbation theory, the effective spin Hamiltonian within the PS manifold becomes (see Appendix \ref{sec:pert_theory})
\begin{align}
  H_{\t{spin}}^{\t{eff}}
  = \f{\p{J\phi}^2}{\p{N-1}U}
  \times \sp{S_\z^2 - N\sum_q s_{\z,q}^2},
\end{align}
whose mean-field version is
\begin{align}
  H_{\t{MFT}}^{\t{eff}} = -\f{\p{J\phi}^2}{U} \sum_q s_{\z,q}^2,
\end{align}
where we have used the fact that the axial magnetizations $\bk{s_{\z,q}}=\frac1N\bk{S_\z}$ within the PS manifold, and $\bk{S_\z}=0$ is conserved by $H_{\t{spin}}^{\t{eff}}$.
The weak-field effective Hamiltonian preserves permutational symmetry, so $\bbk{\ss}_\MF\to1$ as $J\phi/U\to0^+$.
Moreover, permutational symmetry together with the fact that $\bk{S_\z}=\bk{S_\y}=0$ are conserved\footnote{Both the initial state $\ket\x^{\otimes N}$ and the spin Hamiltonian $H_{\t{spin}}$ are invariant under the action of $U_\x=e^{-\i\pi S_\x}\sum_{q>0}\v s_q\c\v s_{-q}$, so $\bk{S_\y}=\bk{U_\x^\dag S_\y U_\x}=-\bk{S_\y}=0$ at all times.} implies that the magnetization $m_\MF$ is determined by the time-average of $\sigma_\x\equiv s_\x/s$ for a single spin:
\begin{align}
  \lim_{J\phi/U\to0^+} m_\MF
  = \abs{\lim_{T\to\infty} \f1T \int_0^T \dd\tau
  \bk{\x|\sigma_\x\p{\tau}|\x}},
\end{align}
where
\begin{align}
  \sigma_\x\p{\tau} = e^{\i\tau s_\z^2} \sigma_\x e^{-\i\tau s_\z^2}.
\end{align}
By mapping the Hilbert space of an $n$-level spin onto the PS (Dicke) manifold of $n-1$ spin-$\frac12$ particles, we can adapt exact analytical results for the Ising model \cite{foss-feig2013nonequilibrium}\footnote{See Appendix K of Ref.~\cite{perlin2020shorttime} for a simpler adaptation of the analytics in Ref.~\cite{foss-feig2013nonequilibrium} to the one-axis twisting model $H_{\t{OAT}}=\chi s_\z^2$.} to find that
\begin{align}
  \bk{\x|\sigma_\x\p{\tau}|\x} = \p{\cos\tau}^{n-2},
\end{align}
so for even $n$
\begin{align}
  \lim_{J\phi/U\to0^+} m_\MF
  = \f1{2\pi} \int_0^{2\pi} \dd\tau \p{\cos\tau}^{n-2}
  = \gamma\p{\f{n}{2}}.
\end{align}

\subsection{Multi-cat initial states}

We now consider the same setup as above, but with the initial ``multi-cat'' states $\ket{\x\x}^{\otimes N}$ and $\ket{\x\x_\i}^{\otimes N}$, where
\begin{align}
  \ket{\x\x} &\equiv \f{\ket{\x} + \ket{-\x}}{\sqrt{2}}
  \simeq e^{-\i\frac{\pi}{2}s_\y^2} \ket{-s}, \\
  \ket{\x\x_\i} &\equiv \f{\ket{\x} + \p{-1}^s \ket{-\x}}{\sqrt{2}}
  \simeq e^{-\i\frac{\pi}{2}s_\x} \ket{\x\x}.
\end{align}
Here $\ket{-\x}\equiv e^{-\i\frac{\pi}{2}s_\y}\ket{-s}$, and $\simeq$ denotes equality up to a global phase.
